Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-11 paper.Rmd
add color to ICEWS models in figures 1 and 2

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
---

@@

library(bookdown)
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```
# Introduction
@blair:sambanis:2020 (hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory, forecasting is more accurate than has previously been possible. Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article. We find that these claims are false. Their theory-based escalation model does not do better than the alternatives that they examine.Actually, it does worse. The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure. The performance results they report are based on smoothed performance curves, not the original unsmoothed curves. Two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

@@

\begin{figure}
\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
\centering
\includegraphics[width=.8\linewidth]{figures/figure1-replicated.png}
\end{figure}

@@

@@

model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
                  "PITF Only")
benefit %>%
  mutate(model = factor(model, levels = rev(model_levels))) %>%
  ggplot(aes(y = model)) + 
  #facet_wrap(~ horizon, scales = "free_x") +
  geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
  labs(x = "AUC(smoothed ROC) - AUC(ROC)",
       y = "") +
  theme_bw() +
  geom_text(data = ann_text, aes(x = diff, label = label), col = "black") +
  geom_vline(xintercept = 0, linetype = 3) +
  theme(axis.text.y = element_text(size = 12))
```

>>>>

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
---

@@

library(bookdown)
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```
# Introduction
Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory, forecasting is more accurate than has previously been possible. Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article. We find that these claims are false. Their theory-based escalation model does not do better than the alternatives that they examine. Actually, it does worse. The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure. The performance results they report are based on smoothed performance curves, not the original unsmoothed curves. Two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

@@

\begin{figure}
\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
\centering
\includegraphics[width=.95\linewidth]{figures/figure1-replicated.png}
\end{figure}

@@

cols <- c(
  Escalation = "#e41a1c",
  Goldstein = "#ff7f00",
  Quad = "#984ea3",
  CAMEO = "#377eb8",
  Average = "#4daf4a",
  `With PITF Predictors` = "gray50", 
  `Weighted by PITF` = "gray50", 
  `PITF Split Population` = "gray50",
  `PITF Only` = "gray50"
)

@@

model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
                  "PITF Only")
benefit %>%
  mutate(model = factor(model, levels = rev(model_levels))) %>%
  ggplot(aes(y = model)) + 
  #facet_wrap(~ horizon, scales = "free_x") +
  geom_point(aes(x = diff, color = model), alpha = 0.5, size = 2) +
  scale_color_manual(guide = FALSE, values = cols) +
  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
       y = "") +
  theme_bw() +
  geom_text(data = ann_text, aes(x = diff, label = label), col = "black") +
  geom_vline(xintercept = 0, linetype = 3) +
  theme(axis.text.y = element_text(size = 12))
```
